Gaia identifies VFTS682 as the most massive, dynamically ejected runaway star known to date

### 2018-05-11 VFTS682.tex
major revision

\documentclass[apjl,twocolumn]{emulateapj}

\usepackage{graphicx}
\usepackage[varg]{txfonts}

\usepackage[pagebackref,breaklinks,colorlinks,citecolor=blue]{hyperref}        
\renewcommand*{\backref}[1]{[#1]}  % gives neat back references in the pdf
%\usepackage[%draft, 
%colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue]{hyperref}
\usepackage{aas_macros}
%\usepackage{amsmath} % clashes with emulateapjj for some reaons
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{comment}
\usepackage{multirow}
%\usepackage{chngpage}
%\usepackage{lscape}
\usepackage{url}
\newcommand{\todo}[1]{{\large $\blacksquare$~\textbf{\color{red}[#1]}}~$\blacksquare$}
\newcommand{\udef}{\stackrel{\mathrm{def}}{=}}
% for tree diagram
%\usepackage{forest}
%\usepackage{tikz-qtree}
%\usetikzlibrary{shadows,trees}

@@

% \defcitealias{tauris:98}{TT98}
% \defcitealias{ramirez-agudelo:15}{R-A15}

@@

\title{ \SdM{Gaia identifies VFTS682 as the most massive, dynamically ejected runaway star known to date}}

\author{M.~Renzo$^{1}$, S.~E.~de~Mink$^{1}$, \SdM{D.J. Lennon$^{2}$ et al.} \todo{... add coauthors...}%\and
  % S.~Justham\inst{2,3} \and A.~de~Koter\inst{1} \todo{etc..order to be decided}
  .}

\affil{$^{1}$Astronomical Institute Anton Pannekoek, University of
    Amsterdam, 1098 XH Amsterdam, The Netherlands \\
    $^{2}$ ESA, European Space Astronomy Centre, Apdo. de Correos 78, E-28691 Villanueva de la Ca\~nada, Madrid, Spain \\
    $^{3}$ ... 
  % \and{School of Astronomy \& Space Science, University of the Chinese
  %   Academy of Sciences, Beijing 100012, China}
  % \and{National Astronomical Observatories, Chinese Academy of
  %   Sciences, Beijing 100012, China}
}

\thanks{M.~Renzo, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}}
\date{}
\begin{abstract}
\SdM{ The formation and fates of the most massive stars remains one of the most elusive questions in stellar astrophysics. 
%
Previous work spectroscopic work identified VFTS682 as one of the most massive stars known. It resides in the field of the 30 Dor region in the Large Magellanic Cloud at a projected distance of 32 pc of the star cluster R136. The lack of a detected offset in radial velocity derived from the spectra, lead to speculations whether such extreme stars can form in relative isolation. 
%
We use Gaia's unprecedented proper motion measurements to investigate the kinematics of this object. We show that VFTS 682 is a bonafide runaway star moving with a relative velocity of about $60\kms$ away from the star cluster R136.  The inferred age, distance, and the magnitude and direction of the velocity are consistent a dynamical ejection scenario from the cluster. 
%
If correct, this would make VFTS 682 the most massive dynamically ejected runaway star known to date.  It proves that star clusters are capable of ejecting their most massive members. We discuss the implications for the existence of stars with masses well above the canonical upper mass limit inside the cluster, the internal dynamics in dense star clusters and the final fate of this object. }

@@

\SdM{

How do massive stars form is one of the major longstanding questions in astrophysics
\citep[e.g.,][]{zinnecker:07}. Obtaining clues from observations has been challenging, because massive stars are intrinsically rare, 
%\citep[e.g.,][]{salpeter:55,kroupa:01, schneider:18}, 
evolve fast, typically reside in dense groups and they remain enshrouded in their parent cloud during the formation
process.  Major progress has been made on the theoretical side,  \citep[e.g.][]{Kuiper+2015,Rosen+2016}, but the simulations are computationally very expensive. They can typically not yet resolve the length scales that are of interest to understand the 
high level of multiplicity among massive stars  \citep[][]{sana:12,sana:17}.  Understanding massive star formation, and its possible dependence on environment and metallicity, is crucial for understanding the role massive stars play within their host galaxies as sources of feedback, but also for understanding the transients that mark their death and the compact remnants they leave behind.  This is therefore a key question for the present and upcoming transient surveys e.g., LSST and LIGO/Virgo,
%\citep[][]{}\todo{ref} . Blackgem is still hypothetical and a very tiny project compared to the others.  
which  will reveal transients associated to massive stars
evolution and death.

In has been conjectured that most, if not all, stars form in clusters \cite{lada:03}, where massive stars are thought to  form preferentially in the dense cores of clusters. In this picture, field stars are primarily the result of the dissolution of dense groups. 
 However, a significant population of massive stars has been  in relative isolation,  far from dense clusters or OB associations. The origin of this population still remains debated \citep{Lamb+2016}.   One hypothesis is that they  formed in the field, but this poses a challenge for the theories of star formation.  The alternative hypothesis is that these massive stars  formed in clusters, and then been ejected from their birth locations.  Such ejections may result from   \citep[e,g,][]{poveda:67} or from the dsiruption of binary systems by the supernova of the companion star  \citep[][]{zwicky:57, blaauw:61}.

 A very interesting contribution to the debate whether or not massive stars an form in relative isolation was presented in recent years by  \cite{bestenlehner:11}, who discuss the case of the very massive star VFTS682.   This star is located in the field of the 30 Doradus region in the Large Magellanic Cloud (LMC) and was studied as part of the multi-epoch spectroscopic VLT-FLAMES Tarantula Survey (VFTS) of massive stars \citep{Evans+2011}.  It's spectral type WNh5,  inferred present-day mass of   $138^{+xx}_{-xx}\,M_\odot$

 an inferred initial mass of $150^{+28.7}_{-17.4}\,M_\odot$  and a mass loss rate of  $\sim10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$  \citep[][]{bestenlehner:11,schneider:18} it belongs to the most extreme objects in the region.  It is reminiscent of the very  massive stars, a.k.a. ``monster stars'', identified by  \citet{Crowther+2010} and \citet{Crowther+2016}.

Most remarkable is the location of  VFTS682, at a projected distance of $\sim$$29$\,pc from the core of the central ionizing star cluster R136.   \citet{bestenlehner:11} discuss two possibilities for the  origin of VFTS 682. Quoting directly from their abstract:  "either (i) the star either formed in situ, which would have profound implications for the formation mechanism of massive stars, or (ii) VFTS 682 is a slow runaway star that originated from the dense cluster R136, which would make it the most massive runaway known to date."

In this study we report on our analysis of Gaia data that was was part of the second data release (DR2) \cite[][]{gaia:16,brown:18}.  We combine the radial velocity measurements from the VFTS survey \citep[][]{evans:11} with the proper motion from Gaia DR2 to reconstruct the three-dimensional velocity of VFTS682, and test the hypothesis that this star was ejected from R136.

Our results indicate that VFTS 682 is a bona fide runaway star.  The direction an magnitude of the velocity vector is fully consistent with dynamical ejection from R136.  Its inferred age is consistent with the inferred travel time. This means that the hypothesis that VFTS 682 is formed in relative isolation is rejected.  This makes VFTS 682, with an inferred present day mass well above a hundred solar masses, the most massive runaway star known to date. 
}

\todo{(1. Make sure we cite and disucss fujii:11, banerjee:12 in discussion. 2. Make link to dynamical interaction proposed to form BHBH systems. }

\section{Observations}
\label{sec:sample}
\SdM{
In this section we summarize the remarkable properties derived for VFTS682 \Secref{data:vfts683} in earlier studies and we discuss how we Gaia data in 
\Secref{data:gaia}.

\subsection{ Overview of the stellar parameters of VFTS682 derived in previous studies \label{data:vfts683}}

The star VFTS682  \citep{Evans+2011}, located at  RA 05:38:55.510  Dec -69:04:26.72 \todo{format properly}
was originally classified as a young stellar object \citep{Gruendl+2009} based on its mid-infrared excess.  \citet{Evans+2011} classified the object a Wolf-Rayet star using spectroscopic observations taken a part of the VLT-FLAMES Tarantula survey.  The spectra available covered $\lambda$4000--7000 and were taken at multiple epochs. \citep{Bestenlehner+2011} studied this object in more detail used and looked for radial velocity variations. This allowed them to exclude the presence of a close companion with high confidence.

\citep{Bestenlehner+2011} further analyzed the spectra using the non-LTE model atmosphere code CMFGEN \citep{Hillier+1998} to derive the stellar parameters and surface abundances. They derive a find a peculiar extinction ($R_V \sim 4.7$), leading to a surprisingly high luminosity $\log_{10} L (\rm{L_\odot}) =  6.5\pm0.2$. Bayesian analysis with the Bonsai code \citep{Schneider+2017} against  evolutionary models by \citet{Brott+2011} and \citet{Kohler+2015} leads to an estimated present-day mass of  $138^{+28}_ {-16}\Msun$, an age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial masse of $150^{+29}_{-17}\Msun$.  This places around the ``canonical upper limit'' of $150\Msun$  by \citet{Figer2005}.

The star is very similar to R136a3, located inside the star cluster R136, \citep{Crowther+2010}  for which \citet{Crowther+2016} report a current mass estimate of $180^{30}_{-30\Msun}$. The R136 cluster hosts two further very massive WN5h stars R1361 and R136a2, whose estimated current masses are even higher.   However, VFTS682 stands out by its location a projected offset of 32\,pc from the center of the cluster.

Further worth noticing is the variability of the star.  \citep{Bestenlehner+2011} discuss the Optical Gravitational Lensing Experiment (OGLE-III) light curves \citep{Udalski+2008} and show a variability in the V-band of the 10\% level on a timescale of years.  The comment that this is unusual for Wolf-Rayet and more reminiscent of Luminous Blue Variables.

\todo{Report mass loss rate.  Report LOS velocity.}

%\citep{Parker1993}

\subsection{Kinematic data from Gaia DR2  \label{data:gaia}}

The Gaia Data Release 2, which became available on 25 April 2018, provides the five-parameter astrometric solution (positions, parallaxes, and proper motions) for more than a bilion sources \citep{Gaia-Collaboration+2018}.  The median uncertainty in the proper motion components the corresponding uncertainties are $0.05\,\masyr$ for bright sources  ($G<14$ mag) and $0.2\masyr$ at $G=17\,\mathrm{mag}$ \citep{Lindegren+2018}.  The uncertainties quoted above to $xx\kms$ and $yy\kms$ respectively.  This unprecedented accuracy allows the identification of bright runaway stars at the distance to the LMC, 50\,kpc.

VFTS682 is identified with the gaia source} with id 4657685637907503744 in the Gaia DR2
catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}. Gaia reports a G-band magnitude is 15.65. The star has a
\texttt{visibility\_period} = 17, which counts how many observations have
been used to reconstruct its astrometric solution
\citep[][]{lindengren:18}, and the reported
\texttt{astrometric\_excess\_noise} = 0. These values suggest that the Gaia
data for VFTS682 are trustworthy. However, the effective temperature
reported in Gaia DR2 is one order of magnitude lower than what found by
\cite{bestenlehner:11}, and the best fit parallax of this star is
negative. We do not use the effective temperature of the star anywhere
in this study, and we attribute the unphysical value of the parallax
to the large distance to the LMC. Our main findings do not rely on the
parallax nor the effective temperature values reported in the Gaia DR2
catalog.

%% should I really mention the line of sight velocity? Not really
%% using it...

We retrieve for VFTS682 the position in right ascension (RA) and declination (DEC)
in the IGCS frame \cite[][]{brown:18}, its
proper motion components ($\mu_\mathrm{RA}$, and $\mu_\mathrm{DEC}$,
respectively). For the radial velocity of VFTS682 and of the 30 Doradus
region as a whole, we instead use the VFTS data
as quoted in \cite{bestenlehner:11}. \Tabref{tab:vfts682} lists the values adopted throughout
this work for each of these quantities.

\begin{table}
\begin{center}
    \caption{Astrometric parameters for VFTS682. }
%
  \begin{tabular}{l|c|c}
  \tableline
\tableline
    Parameter & Value & Source\\ 
    \tableline
    RA \hfill[degree] &  \phantom{-}84.73 $\pm$  0.03 & \multirow{4}{*}{Gaia DR2}\\
    DEC \hfill [degree] & -69.07 $\pm$  0.05  & \\
    $\mu_\mathrm{RA}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}1.84 $\pm$ 0.07 & \\
    $\mu_\mathrm{DEC}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}0.78 $\pm$ 0.08& \\
    $\delta v_\mathrm{rad}$  \hfill[$\kms$] & \phantom{0}-45 $\pm$ 25 & \cite{bestenlehner:11}\\
\tableline
  \end{tabular}
  \tablecomments{The peculiar radial
    velocity $\delta v_\mathrm{rad}$ is obtained as the difference
    between the average radial velocity of the 30 Doradus region
    ($270\pm10\kms$) minus the radial velocity measured from the HeII $\lambda4686$
    line for VFTS682 ($315\pm15\kms$).}
    \end{center}
  \label{tab:vfts682}
\end{table}

%% Any table notes must follow the \end{tabular} command.

%%% change \tablenotetext{a} -> \tablenotetext{1} etc. to omit LaTeX errors
%%% Karol Kozio? for ShareLaTeX

We define a local standard frame \SdM{of reference} to derive the peculiar velocity
of VFTS682 by selecting from the Gaia DR2 catalog a sample of nearby
stars, following closely the approach of \cite{lennon:18}.
We select all the stars in a target of 0.2 degrees around R136
(NGC2070) fulfilling the following criteria. First, we require G-band
magnitude brighter than 17, correspondingly roughly to the
completeness level of the VFTS survey \citep[here we implicitly assume
G$\sim$V,][]{evans:11}. Then we require \texttt{visibility\_period} $\geq$ 5,
\texttt{astrometric\_excess\_noise} < 1, the errors on the proper
motion components to be smaller than 0.1\,$\mathrm{mas\ yr^{-1}}$,
and the proper motion components themselves to be smaller than
2\,$\mathrm{mas\ yr^{-1}}$ in absolute value. At the distance to the
LMC, $1\mathrm{mas\ yr^{-1}}\simeq250\,\kms$ \citep[e.g.,][]{lennon:18}, so the cut on the values
of the proper motions removes stars that would have projected
tangential velocities in excess of $\sim$500\,$\kms$, which are most
likely to be foreground stars. We checked that the additional
requirement of having parallaxes smaller than $2\,\mathrm{mas}$ does
not reduces further our sample.

@@

Using the 651 stars selected as described in \Secref{sec:sample} (a
subset is shown in blue in \Figref{fig:main}), we find averaged proper motion components of
$\langle\mu_\mathrm{RA}\rangle = 1.683\pm0.002\,\mathrm{mas\ yr^{-1}}$ and
$\langle\mu_\mathrm{DEC}\rangle =
0.672\pm0.003\,\mathrm{mas\ yr^{-1}}$. We note that these values are
in good agreement with what found by \cite{lennon:18}. Subtracting these values from the
proper motions of VFTS682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\mu_\mathrm{RA} = 0.16\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\mu_\mathrm{DEC} =
0.11\pm 0.08\,\mathrm{mas\ yr^{-1}}$. We note that the error budget is
dominated by the errors on the proper motion components of VFTS682.

@@

The red arrow in \Figref{fig:main} shows the proper motion of VFTS682
relative to the region, and the lighter red arrows show the possible
range of directions within the uncertainties in the measured proper
motion. It is clear that the most likely origin of the star is R136,
as predicted by \cite{fujii:11, banerjee:12}.

@@

\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{v_\parallel} =
  \frac{29\,\mathrm{pc}}{46\,\kms} \simeq 0.63\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel =29$\,pc is the projected distance from VFTS682 to
the core of R136 \citep[][]{bestenlehner:11}, $v_\parallel \equiv \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2} =46\pm
17\,\kms$ is the speed of the star relative to the cluster projected on the sky, obtained
using the relative proper motion components calculated in
\Secref{sec:runaway}, and we use the approximation $1 \kms \simeq 1
\mathrm{pc \ Myr^{-1}}$.

The kinematic age $\tau_\mathrm{kin}$ is smaller than the apparent age
of the star age $1.0\pm 0.2$\,Myr from \cite{schneider:18}, which
corroborates the idea that the star is the result of a dynamical
ejection.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figures/main_plot}  
  \caption{The red cross indicates the position of VFTS682. Blue stars
    are those we use to define the generic ``surroundings'', while the
    green dots indicate the stars we use to probe the core of R136,
    magnified in the inset. The red arrow shows the direction of the
    proper motion of VFTS682 relative to R136, and the gray shade
    indicates the uncertainty in the direction. \todo{load color picture on background}}
  \label{fig:main}
\end{figure}

@@

Based on our results, we can claim that VFTS682 is the most massive
runaway known to date, with a peculiar three-dimensional spatial
velocity of $\sim$$60\,\kms$. This means that isolated star formation is
\emph{not} required to explain this star. Its proper motion suggests that it was ejected from the cluster R136
$\sim$$0.6$\,Myr ago. Because of the exceptionally large mass
of this star, this raises the question of which stars must populate
the core of the cluster.

>>>>

\documentclass[apjl,twocolumn]{emulateapj}
\usepackage{graphicx}
\usepackage[varg]{txfonts}
\usepackage[pagebackref,breaklinks,colorlinks,citecolor=blue]{hyperref}        
\renewcommand*{\backref}[1]{[#1]}  % gives neat back references in the pdf
%\usepackage[%draft, 
%colorlinks,citecolor=blue,linkcolor=blue,urlcolor=blue]{hyperref}
\usepackage{aas_macros}
%\usepackage{amsmath} % clashes with emulateapjj for some reaons
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{comment}
\usepackage{multirow}
%\usepackage{chngpage}
%\usepackage{lscape}
\usepackage{url}
\newcommand{\todo}[1]{{\large $\blacksquare$~\textbf{\color{red}[#1]}}~$\blacksquare$}
\newcommand{\udef}{\stackrel{\mathrm{def}}{=}}
% for tree diagram
%\usepackage{forest}
%\usepackage{tikz-qtree}
%\usetikzlibrary{shadows,trees}

@@

@@

\title{Gaia identifies VFTS682 as the most massive, dynamically ejected runaway star known to date}

\author{M.~Renzo$^{1}$, S.~E.~de~Mink$^{1}$, D.~J.~Lennon$^{2}$,
  S.~Justham$^{3,4}$, J.~Vink$^{5}$, A.~de~Koter$^{1}$ \todo{... add coauthors...} }

\affil{$^{1}$Astronomical Institute Anton Pannekoek, University of
    Amsterdam, 1098 XH Amsterdam, The Netherlands \\
    $^{2}$ ESA, European Space Astronomy Centre, Apdo. de Correos 78, E-28691 Villanueva de la Ca\~nada, Madrid, Spain \\
    $^{3}$ 
  % \and{School of Astronomy \& Space Science, University of the Chinese
  %   Academy of Sciences, Beijing 100012, China}
  % \and{National Astronomical Observatories, Chinese Academy of
  %   Sciences, Beijing 100012, China}
}

\thanks{M.~Renzo, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}}
\date{}
\begin{abstract}
The formation and fates of the most massive stars remains one of the most elusive questions in stellar astrophysics. 
%
Previous spectroscopic studies identified VFTS682 as one of the most massive stars known. It resides in the field of the 30 Dor region in the Large Magellanic Cloud at a projected distance of 29\,pc of the star cluster R136. The lack of a detected offset in radial velocity derived from the spectra, lead to speculations whether such extreme stars can form in relative isolation. 
%
We use Gaia's unprecedented proper motion measurements to investigate the kinematics of this object. We show that VFTS 682 is a bonafide runaway star moving with a relative velocity of about $60\kms$ away from the star cluster R136.  The inferred age, distance, and the magnitude and direction of the velocity are consistent a dynamical ejection scenario from the cluster. 
%
If correct, this would make VFTS 682 the most massive dynamically ejected runaway star known to date.  It proves that star clusters are capable of ejecting their most massive members. We discuss the implications for the existence of stars with masses well above the canonical upper mass limit inside the cluster, the internal dynamics in dense star clusters and the final fate of this object.

@@

How do massive stars form is one of the major longstanding questions in astrophysics
\citep[e.g.,][]{zinnecker:07}. Obtaining clues from observations has been challenging, because massive stars are intrinsically rare, 
%\citep[e.g.,][]{salpeter:55,kroupa:01, schneider:18}, 
evolve fast, typically reside in dense groups and they remain enshrouded in their parent cloud during the formation
process.  Major progress has been made on the theoretical side,
\citep[e.g.][]{kuiper:15,rosen:16}, but simulations cannot yet resolve
the smallest length scales that are of interest to understand the 
high level of multiplicity among massive stars  \citep[][]{sana:12,sana:17}.  Understanding massive star formation, and its possible dependence on environment and metallicity, is crucial for understanding the role massive stars play within their host galaxies as sources of feedback, but also for understanding the transients that mark their death and the compact remnants they leave behind.  This is therefore a key question for the present and upcoming transient surveys e.g., LSST and LIGO/Virgo,
%\citep[][]{}\todo{ref} . Blackgem is still hypothetical and a very tiny project compared to the others.  
which  will reveal transients associated to massive stars
evolution and death.

In has been proposed that most, if not all, stars form in clusters \citep{lada:03}, where massive stars are thought to  form preferentially in the dense cores of clusters. In this picture, field stars are primarily the result of the dissolution of dense groups. 
However, a significant population of massive stars exists in relative
isolation,  far from dense clusters or OB associations. The origin of
this population still remains debated \citep{lamb:16}.   One
hypothesis is that they  formed in the field, but this poses a
challenge for the theories of star formation.  The alternative
hypothesis is that these massive stars  formed in clusters, and then
been ejected from their birth locations.  Such ejections may result
from dynamical interactions \citep[e.g.,][]{poveda:67} or from the disruption of binary systems by the core-collapse of the companion star \citep[e.g.,][]{zwicky:57, blaauw:61}.

An interesting contribution to the debate on whether or not massive
stars an form in relative isolation was presented by
\cite{bestenlehner:11}, who discuss the case of the very massive star
VFTS682.   This star is located in the field of the 30 Doradus region
in the Large Magellanic Cloud (LMC) and was studied as part of the
multi-epoch spectroscopic VLT-FLAMES Tarantula Survey (VFTS) of
massive stars \citep{evans:11}.  Its spectral type is WNh5 and its 
present-day mass is $137.8^{+27.5}_{-15.9}\,M_\odot$,
\citep{schneider:18}. These correspond to an inferred initial mass of
$150.0^{+28.7}_{-17.4}\,M_\odot$ and a mass loss rate of
$\sim10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$
\citep[][]{bestenlehner:11,schneider:18}, which make it one of the
most extreme objects in the region.  It is reminiscent of the very
massive stars, a.k.a. ``monster stars'', identified by
\citet{crowther:10, crowther:16} in the core of the
R136 cluster.

Most remarkable is the location of VFTS682, at a projected distance of
$\sim$$29$\,pc from the core of the star cluster
R136. \citet{bestenlehner:11} discuss two possibilities for the
origin of VFTS 682. Quoting directly from their abstract: "either (i) the star either formed in situ, which would have profound implications for the formation mechanism of massive stars, or (ii) VFTS 682 is a slow runaway star that originated from the dense cluster R136, which would make it the most massive runaway known to date."

In this study we report on our analysis the data from the second Gaia data release (DR2) \cite[][]{gaia:16,brown:18}.  We combine the radial velocity measurements from the VFTS survey \citep[][]{evans:11} with the proper motion from Gaia DR2 to reconstruct the three-dimensional velocity of VFTS682, and test the hypothesis that this star was ejected from R136.

Our results indicate that VFTS 682 is a bona fide runaway star.  The direction an magnitude of the velocity vector is fully consistent with dynamical ejection from R136.  Its inferred age is consistent with the inferred travel time. This means that the hypothesis that VFTS 682 is formed in relative isolation is rejected.  This makes VFTS 682, with an inferred present day mass well above a hundred solar masses, the most massive runaway star known to date.

\section{Observations}
\label{sec:sample}

In this section we summarize the remarkable properties derived for VFTS682 \todo{?}\Secref{data:vfts683}
in earlier studies and we discuss we Gaia data in \Secref{data:gaia}.

\subsection{ Overview of VFTS682 from previous studies \label{data:vfts683}}

The star VFTS682  \citep{evans:11}, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) -69:04:26.72 in the ICGS frame \citep[][]{brown:18}
was originally classified as a young stellar object \citep{gruendl:09}
based on its mid-infrared excess.  \citet{evans:11} classified the
object as Wolf-Rayet star of spectral type WNh5 using spectra from the
VFTS survey.  The spectra available covered $\lambda$4000--7000 and
were taken at multiple epochs. Using the same dataset,
\citet{bestenlehner:11} excluded the presence of a close companion
with high confidence from the absence of radial velocity variations.

\citet{bestenlehner:11} further analyzed the spectra using the non-LTE
model atmosphere code CMFGEN \citep{hillier:98} to derive the stellar
parameters and surface abundances. They inferred a peculiar extinction
($R_V\simeq4.7$), leading to a surprisingly high luminosity
$\log_{10}(L/L_\odot) =  6.5\pm0.2$. \citep{schneider:18} estimated
the present-day mass of VFTS682 at  $137.8^{+27.5}_ {-15.9}\Msun$, an
apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass
of $150.0^{+28.7}_{-17.4}\Msun$, using the Bayesian analysis tool Bonsai
\citep{schneider:17}, based on the evolutionary models from
\citet{brott:11, kohler:15}. These values place VFTS682 around the ``canonical upper limit'' of $\sim$$150\Msun$  by \citet{figer:05}.

The star is very similar from the spectral point of view to R136a3, located inside the star cluster
R136, \citep{crowther:10}  for which \citet{crowther:16} report a
current mass estimate of $180^{+30}_{-30}\Msun$. The R136 cluster hosts
at least two more very massive WN5h stars R136a1, R136a2 whose
estimated current masses are even higher. However, VFTS682 stands
out by its isolation at a projected offset of 29\,pc from the center
of the cluster \citep{bestenlehner:11}.

Further worth noticing is the variability of the star.
\citep{bestenlehner:11} discuss the Optical Gravitational Lensing
Experiment (OGLE-III) light curves \citep{udalski:08} and show a
variability in the V-band of the $\sim$10\% level on a timescale of years.
The comment that this is unusual for Wolf-Rayet and more reminiscent
of Luminous Blue Variable (LBV) stars.

\todo{Report mass loss rate.  Report LOS velocity.}
\citep{parker:93}

\subsection{New data from Gaia DR2  \label{data:gaia}}

\begin{table}[t]
\begin{center}
    \caption{Astrometric parameters for VFTS682. }
%
  \begin{tabular}{l|c|c}
  \tableline
\tableline
    Parameter & Value & Source\\ 
    \tableline
    RA \hfill[degree] &  \phantom{-}84.73 $\pm$  0.03 & \multirow{4}{*}{Gaia DR2}\\
    DEC \hfill [degree] & -69.07 $\pm$  0.05  & \\
    $\mu_\mathrm{RA}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}1.84 $\pm$ 0.07 & \\
    $\mu_\mathrm{DEC}$  \hfill[$\mathrm{mas\ yr^{-1}}$] & \phantom{-0}0.78 $\pm$ 0.08& \\
    $\delta v_\mathrm{rad}$  \hfill[$\kms$] & \phantom{0}-45 $\pm$ 25 & \cite{bestenlehner:11}\\
\tableline
  \end{tabular}
  \tablecomments{The peculiar radial
    velocity $\delta v_\mathrm{rad}$ is obtained as the difference
    between the average radial velocity of the 30 Doradus region
    ($270\pm10\kms$) minus the radial velocity measured from the HeII $\lambda4686$
    line for VFTS682 ($315\pm15\kms$).}
    \end{center}
  \label{tab:vfts682}
\end{table}

% %% Any table notes must follow the \end{tabular} command.

% %%% change \tablenotetext{a} -> \tablenotetext{1} etc. to omit LaTeX errors
% %%% Karol Kozio? for ShareLaTeX

The Gaia Data Release 2, which became available on 25 April 2018,
provides the five-parameter astrometric solution (positions,
parallaxes, and proper motion components) for more than a billion
sources \citep{brown:18}.%  \todo{why do we care about the
%   median? I'm
%   selecting data with errors smaller than this, and I'm not using
%   fainter stars!I think this is not useful here} The median uncertainty in the proper motion
% components the corresponding uncertainties are $0.05\,\masyr$ for
% bright sources  ($G<14$ mag) and $0.2\masyr$ at $G=17\,\mathrm{mag}$
% \citep{lindegren:18}.  The uncertainties quoted above correspond to $xx\kms$ and $yy\kms$ respectively.  This unprecedented accuracy allows the identification of bright runaway stars at the distance to the LMC, 50\,kpc.

VFTS682 is identified with the id 4657685637907503744 in the Gaia DR2
catalog\footnote{\url{https://vizier.u-strasbg.fr/viz-bin/VizieR-3?-source=I/345/gaia2}}. Gaia
reports a G-band magnitude of 15.65. The star has a
\texttt{visibility\_period} = 17, which counts how many observations have
been used to reconstruct its astrometric solution
\citep[][]{lindengren:18}, and the reported
\texttt{astrometric\_excess\_noise} = 0. These values suggest that the Gaia
data for VFTS682 are trustworthy. However, the effective temperature
reported in Gaia DR2 is one order of magnitude lower than what found by
\cite{bestenlehner:11}, and the best fit parallax of this star is
negative. We do not use the effective temperature of the star anywhere
in this study, and we attribute the unphysical value of the parallax
to the large distance to the LMC. Our main findings do not rely on the
parallax nor the effective temperature values reported in the Gaia DR2
catalog.

We retrieve for VFTS682 the position in RA and DEC
in the IGCS frame \cite[][]{brown:18}, its
proper motion components ($\mu_\mathrm{RA}$, and $\mu_\mathrm{DEC}$,
respectively). For the radial velocity of VFTS682 and of the 30 Doradus
region as a whole, we instead use the VFTS data
as quoted in \cite{bestenlehner:11}. \Tabref{tab:vfts682} lists the values adopted throughout
this work for each of these quantities.

We define a local standard frame of reference to derive the peculiar velocity
of VFTS682 by selecting from the Gaia DR2 catalog a sample of nearby
stars, following closely the approach of \cite{lennon:18}.
We select all the stars in a target of 0.2 degrees around R136
(NGC2070) fulfilling the following criteria. First, we require G-band
magnitude brighter than 17, correspondingly roughly to the
completeness level of the VFTS survey \citep[here we implicitly assume
G$\sim$V,][]{evans:11}. Then we require \texttt{visibility\_period} $\geq$ 5,
\texttt{astrometric\_excess\_noise} $< 1$, the errors on the proper
motion components to be smaller than 0.1\,$\mathrm{mas\ yr^{-1}}$,
and the proper motion components themselves to be smaller than
2\,$\mathrm{mas\ yr^{-1}}$ in absolute value. At the distance to the
LMC, $1\mathrm{mas\ yr^{-1}}\simeq250\,\kms$ \citep[e.g.,][]{lennon:18}, so the cut on the values
of the proper motions removes stars that would have projected
tangential velocities in excess of $\sim$500\,$\kms$, which are most
likely to be foreground stars. We checked that the additional
requirement of having parallaxes smaller than $2\,\mathrm{mas}$ does
not reduces further our sample.

@@

Using the 651 stars selected as described in \Secref{data:gaia} (a
subset is shown in blue in \Figref{fig:main}), we find averaged proper motion components of
$\langle\mu_\mathrm{RA}\rangle = 1.683\pm0.002\,\mathrm{mas\ yr^{-1}}$ and
$\langle\mu_\mathrm{DEC}\rangle =
0.672\pm0.003\,\mathrm{mas\ yr^{-1}}$. We note that these values are
in good agreement with what found by \cite{lennon:18}. Subtracting these values from the
proper motions of VFTS682 (see \Tabref{tab:vfts682}), we obtain the
components of proper motion of the star relative to the surrounding region
$\mu_\mathrm{RA} = 0.16\pm 0.07\,\mathrm{mas\ yr^{-1}}$ and $\mu_\mathrm{DEC} =
0.11\pm 0.08\,\mathrm{mas\ yr^{-1}}$. We note that the error budget is
dominated by the errors on the proper motion components of VFTS682.

@@

The red arrow in \Figref{fig:main} shows the proper motion of VFTS682
relative to the region, and the lighter red arrows show the possible
range of directions within the uncertainties in the measured proper
motion. It is clear that the most likely origin of the star is R136,
as predicted from numerical N-body simulations by \cite{fujii:11, banerjee:12}.

@@

\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{v_\parallel} \simeq
  \frac{29\,\mathrm{pc}}{46\,\kms} \simeq 0.63\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel =29$\,pc is the projected distance from VFTS682 to
the core of R136 \citep[][]{bestenlehner:11}, $v_\parallel \equiv \sqrt{\left(\delta v_\mathrm{RA}\right)^2
    +\left(\delta v_\mathrm{DEC}\right)^2} =46\pm
17\,\kms$ is the speed of the star relative to the cluster projected on the sky, obtained
using the relative proper motion components calculated in
\Secref{sec:runaway}, and we use the approximation $1 \kms \simeq 1
\mathrm{pc \ Myr^{-1}}$.

The kinematic age $\tau_\mathrm{kin}$ is smaller than the apparent age
of the star age of $1.0\pm 0.2$\,Myr from \cite{schneider:18}, which
corroborates the idea that the star is the result of a dynamical
ejection.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figures/main_plot}  
  \caption{The red solid arrow indicates the proper motion of VFTS682
    relative to the region, starting from the present day position of
    the star. The semi-transparent arrows indicate the possible
    directions of projected motion within the errors, and are extended
    backwards to illustrate the uncertainty on the origin of the
    star. The blue stars
    are a subset of those we use to define the local rest frame. \todo{load color picture on background}}
  \label{fig:main}
\end{figure}

@@

Based on our results, we can claim that VFTS682 is the most massive
runaway known to date, with a peculiar three-dimensional speed of $64\pm21\,\kms$. This means that isolated star formation is
\emph{not} required to explain this star. Its proper motion suggests that it was ejected from the cluster R136
$\sim$$0.6$\,Myr ago. Because of the exceptionally large mass
of this star, this raises the question of which stars must populate
the core of the cluster.
